Estimation of Intracortical Connectivity in a Dynamical Neural Field Model

### 2010-03-16 ltx/BrainIDE.tex
updates to paper from meeting on 16/03/2010

\section{Neural Field Model}
In this section, we describe a variant of the  Wilson and Cowan neural field model~\cite{Wilson1973} that will be used to create our intracortical connectivity estimator. This model is descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour. It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology.
The model relates the average number of action potentials $g(r,t)$ arriving at position $r$ to the local post synaptic membrane voltage $v(r,t)$. The post-synaptic potentials generated at a neuronal population at location $r$ by action potentials arriving from all other connected populations at locations $r'$ can be described by
\begin{equation}\label{SpikesToPotential}
v\left( {r,t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {r,t'} \right)dt'}.
\end{equation}
The post-synaptic response kernel $h(t)$ is described by
\begin{equation}\label{SynapticRespKernel}
h(t) = \left\{ {\begin{array}{*{20}{c}}
   {\exp ( - \alpha t)} & {t \ge 0}  \\
   0 & {t < 0}  \\
\end{array}} \right.
\end{equation}
where $\alpha=\tau^{-1}$ and $\tau$ is the synaptic time constant. The synaptic response kernel can be more elaborate, using rise and fall time constants, however, this simple form serves our purposes. Nonlocal interactions between cortical populations are described by	
\begin{equation}\label{RateBasedInteractions}
g\left( r,t \right) = \int_\Omega  {w\left( r,r' \right)f\left( v\left( r',\bar t \right) \right)dr'} + e\left(r,t\right)
\end{equation}
where $\bar t = t - d\left| {r - r'} \right|$, $d$ is the propagation delay of action potentials, $f$ is spiking rate of populations $r'$, $w$ is the spatial connectivity kernel, and $\Omega$ is the spatial domain, representing a cortical sheet or surface. The term $de(r,t)$ can be considered the action potentials generated by unmodeled inputs and spontaneous background activity and is a Wiener process with mean $\mu$ incremental covariance $\Sigma_e(r)dt$.

The average firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function
\begin{equation}\label{ActivationFunction}
f\left( {{v_t}\left( {r'} \right)} \right) = \frac{{{\nu _0}}}{{1 + \exp \left( {\beta \left( {{v_0}\left( r \right) - {v}\left(r',\bar{t} \right)} \right)} \right)}}.
\end{equation}

Even though we will model  $v_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality when constructing the state-space representation in the next section. The parameters in equation~\ref{ActivationFunction} are detailed in Table 1. By combining Eqs.~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
v\left(r,t\right) = \int\limits_{-\infty}^t h\left(t - t'\right) \left(\int\limits_\Omega  w\left(r,r'\right)f\left( v\left( r',\bar t \right)\right)dr' + e\left(r,t\right)\right) dt'
\end{equation}
The rearrange the model into an intergro-differential we take the Laplace transform of~\ref{FullDoubleIntModel} giving
\begin{equation}
	V(s) = H(s)G(s).
\end{equation}
Since $H(s) = (s+\alpha)^{-1}$ we can write
\begin{equation}
	(s+\alpha)V(s) = G(s).
\end{equation}
% To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
% \begin{equation}\label{GreensFuncDef}
% Lh\left( t \right) = \delta \left( t \right),
% \end{equation}
% where $L=\frac{d}{dt} + \alpha$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function. 
Now taking the inverse Laplace transform provides the final form
\begin{equation}\label{FinalForm1}
dv\left( r,t \right) + \alpha v\left( r,t \right)dt = \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',\bar t \right)} \right)dr'dt}  +de\left( r,t \right).
\end{equation}

\section{Discrete Time and Space Model}
We begin the derivation of the state-space model with the continuous time, continuous space IDE neural field model. 
Spatio-temporal interactions can be described by the continuous space continuous time stochastic differential equation
\begin{equation}\label{differential eq}	
dv\left( r,t \right) =  \left( -\alpha v\left( r,t \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t \right)} \right)dr'} \right) dt  + d e\left( r,t \right).
\end{equation}
To simplify the notation, but without loss of generality, \ref{FinalForm1} can be rewritten as
\begin{equation}\label{differential simplified}
	dv\left( r,t \right) = q\left(v\left(r,t\right) \right) dt  + d e\left( r,t \right)
\end{equation}
where $q\left(v\left(r,t\right)\right)$ is the term insides the brackets in Eq.~\ref{differential eq}. A one-step Euler method was used to approximate the nonlinear stochastic IDE to a discrete time form. 
\begin{equation}\label{euler}
	v\left(r,t\right) =  v(r,t_0) + \int_{t_0}^{t} q\left(v(r,\tau) \right) d\tau  + \int_{t_0}^{t} d e\left(r,\tau\right)
\end{equation}
Using an Euler approximation,~\ref{euler} can be approximated at each sampling time by
\begin{equation}\label{euler approx 1}
	v\left(r,t_{k+1} \right) =  v(r,t_k) + q\left(v(r,t_k)\right) (t_{k+1} - t_k) + (e(r,t_{k+1}) - e(r,t_k))
\end{equation}
which can be rewritten as
\begin{equation}\label{euler approx 2}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta q\left(v(r,t_k)\right) + e'(r,t_k)
\end{equation}
where $e'(r,t_k)$ is a zero mean Gaussian random variable with covariance $\delta\Sigma_e$. (Note that, because of the way the noise is defined in \ref{differential eq} it is not $\delta^2$ but $\delta$). Expanding $q\left(v(r,t_k)\right)$ we have
\begin{equation}\label{discrete time model}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta  \left( -\alpha v\left( r,t_k \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} \right) + e'(r,t_k)
\end{equation}
which can be simplified to
\begin{equation}\label{discrete time model2}
	v\left(r,t_k + \delta \right) =  (1 - \delta \alpha) v(r,t_k) + \delta \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} + e'(r,t_k)
\end{equation}
Now we define the spatial aspect of the model on a regular square $i,j$ grid of neural masses, where the spatial step size $\Delta \mathbf{r}_i = \Delta \mathbf{r}_j = \Delta $ giving
\begin{equation}\label{discrete space}
	v\left(r_{ij},t_k + \delta \right) =  (1 - \delta \alpha)v(r_{ij},t_k) + \delta \Delta^2 \sum_{i=-I}^{i=I}{\sum_{j=-J}^{j=J}{\tilde{w}\left( r_{ij},r_{ij}' \right)f\left( v\left( r_{ij}',t_k \right) \right)} }+ e'(r_{i,j},t_k).
\end{equation}
where $\tilde{w}(r_{ij},r_{ij}')$ discretized connectivity kernel. In Eq~\ref{discrete space} we have used periodic boundary conditions to alleviate problems associated with edge effects. The connectivity kernel is in the form of a modified Mexican hat, and is described by the sum of three weighted Gaussian functions as
\begin{equation}
	\tilde{w}(r_{ij},r_{ij}') = \mathbf{\theta}^T e^{\frac{\left(\left(r_i'-r_i\right)\Delta\right)^2}{(2\mathbf{\sigma})^2} + \frac{\left(\left(r_j'-r_j\right)\Delta\right)^2}{(2\mathbf{\sigma})^2}}
\end{equation}
where
\begin{equation}
 \mathbf{\theta}=\begin{bmatrix} \mathbf{\theta}_{1}\\\mathbf{\theta}_{2}\\\mathbf \theta_{3}\end{bmatrix},\quad \mathbf{\sigma}=\begin{bmatrix}\mathbf{\sigma}_1\\\mathbf{\sigma}_2\\ \mathbf{\sigma}_3\end{bmatrix}.
\end{equation}.
The first Gaussian function (parametrized by $\mathbf{\theta}_1, \mathbf{\sigma}_1$) describes local excitation, the second describes lateral inhibition and the third describes medium range excitation.

To generate data will approximate either local field potential recordings or intracranial EEG we have used the observation equation
\begin{equation}\label{ObservationEquation}
    y(r_n,t_k') = \int_\Omega  m( r_n,r' ) v( r',t_k' ) dr'  + \varepsilon ( r_n,t_k ),
\end{equation}
where $\varepsilon( n )\sim\mathcal{N}(0,\Sigma)$, $n$ indexes the iEEG electrodes, and $m(r_n,r)$ is the observation kernel relating the neural field to measurements accounting for electrodes characteristics (spatial averaging). In Eq.~\ref{ObservationEquation} we have indexed time by $t_k'$, which is a decimated version of $t_k$.

@@

\begin{tabular}{ccc}
\hline
Parameter & Value & Reference\\
\hline
$\alpha$ & $100$ s$^{-1}$ &\\
\hline
$\nu_0$ & $1$ s$^{-1}$ &\\
\hline
$v_0$ & $6$ mV &\\
\hline
$\beta$ & $0.56$ spikes/mV$^{-1}$ & \\
\hline
$\delta$ & $1\times10^{-4}$ s &\\
\hline
$\Delta$ & $1$ &\\
\hline
$\theta_1$ &  &\\
\hline
$\theta_2$ &  &\\
\hline
$\theta_3$ &  &\\
\hline
$\sigma_1$ &  &\\
\hline
$\sigma_2$ &  &\\
\hline
$\sigma_3$ &  &\\
\hline
\end{tabular}

@@

In order to implement standard estimation techniques we propose a decomposition of the field using a set of Gaussian basis functions. Decomposition allows a continuous field to be represented by a finite dimensional state vector. This allows for the application of standard nonlinear, non-Gaussian state estimation methods such as the unscented Kalman filter. The abbreviate the notation we use the subscript $t$ to index the current time step and $t+1$ for next time step. The field decomposition is described by
\begin{equation}\label{DefFieldDecomp}
v_t\left( \mathbf{r} \right) = \phi ^T\left( \mathbf{r} \right) x_t.
\end{equation}
where $\phi(\mathbf{r})$ is a vector of Gaussian basis functions that are scaled by the state vector. The connectivity kernel can all be decomposed into a set of basis functions
\begin{equation}\label{DefKernelDecomp}
\tilde{w}( \mathbf{r} , \mathbf{r'} ) = \theta^T\psi(\mathbf{r} , \mathbf{r'} )
\end{equation}
Will will assume we know the parametric form of the basis functions, where the parameter $\theta$ is unknown. Each connectivity basis function can be considered a layer in the WC model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{discrete space} we get

@@

\begin{equation}\label{DecompModel}
{\phi ^T}\left( \mathbf{r} \right){x_{t+1}} = T_s\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\mathbf{r'}}} \right)f\left( {{\phi ^T}\left( {{\mathbf{r'}}} \right){x_t}} \right)d{\mathbf{r'}}}  - \lambda\phi^T(\mathbf{r})x_t + T_s{e_t}\left( {\mathbf{r}} \right).
\end{equation}
To isolate the state we cross multiply by $\phi(\mathbf{r})$ and integrate over $\mathbf{r}$
\begin{equation}\label{Poo}
\int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}{x_{t+1}}}  = T_s\int_\Omega  {\phi ( \mathbf{r} )\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}}  - \lambda\int_\Omega \phi(\mathbf{r})\phi^T(\mathbf{r})d\mathbf{r} x_t  + T_s\int_\Omega  {\phi \left( {\mathbf{r}} \right)e_t\left( {\mathbf{r}} \right)d{\mathbf{r}}}.
\end{equation}
Now we define
\begin{equation}\label{DefGamma}
\Gamma  = \int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
Substituting equation~\ref{DefGamma} into \ref{Poo} and rearranging gives
\begin{equation}\label{SS Model}
{x_{t+1}} = T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}} - \lambda x_t + T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)e_t\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
We can rearrange to get
\begin{equation}\label{Homogeneous SS Model}
	x_{t+1} = T_s\Gamma ^{ - 1}\int_\Omega  {\phi \left( {\bf{r}} \right)\int_\Omega  {f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)\psi^T \left( {{\bf{r}} - {\bf{r'}}} \right)d{\bf{r'}}} d{\bf{r}}}\theta - \lambda x_t + T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)e_t\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
The disturbance term can be rewritten as 
\begin{equation}
	\Gamma ^{ - 1}\int_\Omega  {\phi \left( \bf{r} \right)e_t\left( \bf{r} \right)d\bf{r}} = \int_\Omega  {\phi \left( \bf{r} \right)\tilde{e}_t\left( \bf{r} \right)d\bf{r}},
\end{equation}
where $\tilde{e}_t(r)~N(0,R'R)$ and $R'$ is the Cholesky decomposition of $\Gamma^{-1}$. This provides the final state-space model
\begin{equation}\label{AbbrevSSModel}
x_{t + 1} = T_sq(x_t)\theta -\lambda x_t + T_s\int_\Omega  {\phi ( \mathbf{r} )\tilde{e}_t( \mathbf{r} )d\mathbf{r}}
\end{equation}
with observation equation
\begin{equation}\label{ObservationEquation}
    y_t( k ) = \int_\Omega  m( k,\mathbf{r} )\phi^T( \mathbf{r} ) d\mathbf{r}x_t  + \varepsilon ( k ),
\end{equation}

\subsection{Discrete Space For Simulation}

In this section we describe a discrete time, discrete space model that we are using for the simulations. Now we define the spatial aspect of the model on a regular square $i,j$ grid of neural masses, where the spatial step size $\Delta \mathbf{r}_i = \Delta \mathbf{r}_j = \Delta \mathbf{r}$ giving
\begin{equation}\label{DiscreteSpaceModelij}
	x_{t+1} = T_s\Gamma ^{ - 1}\sum_i \sum_j  {\phi \left( \mathbf{r} \right)\sum_i\sum_j  {f\left( \phi^T \left( \mathbf{r'} \right) x_t \right)\psi^T \left( \mathbf{r} - \bf{r'} \right)\Delta\mathbf{r'}^2} \Delta\mathbf{r}^2} \theta - \lambda x_t + T_s\sum_i\sum_j  {\phi \left( {\bf{r}} \right)\tilde{e}_t\left( \bf{r} \right)\Delta\mathbf{r}^2}.
\end{equation}
\section{Estimation of the Nonlinear Homogeneous IDE Neural Field Model}

>>>>

\section{Neural Field Model}
In this section, we describe a variant of the  Wilson and Cowan neural field model~\cite{Wilson1973} that will be used to create our intracortical connectivity estimator. This model is descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations, and epileptic behaviour. It is also capable of generating complex patterns of activity such as Turing patterns, spirals, and traveling oscillations. The neural field model is popular due to being parsimonious yet having a strong link with the underlying physiology.
The model relates the average number of action potentials $g(r,t)$ arriving at position $r$ to the local post synaptic membrane voltage $v(r,t)$. The post-synaptic potentials generated at a neuronal population at location $r$ by action potentials arriving from all other connected populations at locations $r'$ can be described by
\begin{equation}\label{SpikesToPotential}
	v\left( {r,t} \right) = \int_{ - \infty }^t {h\left( {t - t'} \right)g\left( {r,t'} \right)dt'}.
\end{equation}
The post-synaptic response kernel $h(t)$ is described by
\begin{equation}\label{SynapticRespKernel}
	h(t) = u(t)\exp{\left(-\alpha t\right)}.
\end{equation}
where $\alpha=\tau^{-1}$ and $\tau$ is the synaptic time constant and $u(t)$ is the Heaviside step function. The synaptic response kernel can be more elaborate, using rise and fall time constants, however, this simple form serves our purposes. Nonlocal interactions between cortical populations are described by	
\begin{equation}\label{RateBasedInteractions}
	g\left( r,t \right) = \int_\Omega  {w\left( r,r' \right)f\left( v\left( r',\bar t \right) \right)dr'}
\end{equation}
where $\bar t = t - d\left| {r - r'} \right|$, $d$ is the propagation delay of action potentials, $f$ is spiking rate of populations $r'$, $w$ is the spatial connectivity kernel, and $\Omega$ is the spatial domain, representing a cortical sheet or surface. The connectivity kernel is a Mexican hat shape described by the sum of weighted Gaussian functions
\begin{equation}
	w\left(r,r'\right) = \theta^T \exp{\left(-\frac{\left(r_x - r_{x_n}\right)^2}{\left(2\sigma_w\right)^2}-\frac{\left(r_y - r_{y_n}\right)^2}{\left(2\sigma_w\right)^2}\right)}
\end{equation}
where
\begin{equation}
	\mathbf{\theta}=\begin{bmatrix} \mathbf{\theta}_{1}\\\mathbf{\theta}_{2}\\\mathbf \theta_{3}\end{bmatrix},\quad 		\mathbf{\sigma}=\begin{bmatrix}\mathbf{\sigma}_1\\\mathbf{\sigma}_2\\ \mathbf{\sigma}_3\end{bmatrix}.
\end{equation}
The average firing rate of the presynaptic neurons is related to the postsynaptic membrane potential by the sigmoidal activation function
\begin{equation}\label{ActivationFunction}
	f\left( v_t\left( r' \right) \right) = \frac{\nu _0\left(r\right)}{1 + \exp \left( \beta \left( v_0\left( r \right) - v\left(r',\bar{t} \right) \right) \right)}.
\end{equation}
Even though we will model $v_0$ and $\nu_0$ as being homogeneous across the cortical sheet we will express it as being spatially dependant allowing for generality at this point. By combining Eqs.~\ref{SpikesToPotential} and \ref{RateBasedInteractions} we get the spatiotemporal model
\begin{equation}\label{FullDoubleIntModel}
	v\left(r,t\right) = \int_{-\infty}^t \int_\Omega  h\left(t - t'\right) w\left(r,r'\right) f\left( v\left( r',\bar t \right)\right)dr'dt'
\end{equation}
% The rearrange the model into an integro-differential we take the Laplace transform of~\ref{FullDoubleIntModel} giving
% \begin{equation}
% 	V(s) = H(s)G(s).
% \end{equation}
% Since $H(s) = (s+\alpha)^{-1}$ we can write
% \begin{equation}
% 	(s+\alpha)V(s) = G(s).
% \end{equation}
To arrive at the final form of the model we shall state the synaptic response kernel as a Green's function
\begin{equation}\label{GreensFuncDef}
	Lh\left( t \right) = \delta \left( t \right),
\end{equation}
where $L=\frac{d}{dt} + \alpha$ is a temporal differential operator and $\delta(t)$ is the Dirac-delta function giving 
\begin{equation}\label{FinalForm1}
	\frac{dv\left( r,t \right)}{dt} + \alpha v\left( r,t \right) = \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',\bar t \right)} \right)dr'}.
\end{equation}

\section{Generating Data}
To generate iEEG/LFP data we proceeded with the assumption of spatial homogeneity for the connectivity and activation function parameters, implying $w(r,r') = w\left(|r-r'|\right)$, $\nu_0(r) = \nu_0$ and $v_0(r) = v_0$. In addition we assume the synaptic transmission delay is negligible such that $\bar{t}=t$. The model was discretized in time  (see Appendix~\ref{Time Discretization}) and a stochastic term was added to drive the system giving
\begin{equation}\label{DiscreteTimeModel}
	v_{t+1}\left(r\right) = \lambda v_t\left(r\right) + \delta \int_\Omega { w\left(|r-r'|\right) f\left(v_t\left(r'\right)\right) dr'} + e_t\left(r\right)
\end{equation}
where $\delta$ is the time step, $\lambda = 1-\alpha\delta$ and $e_t(r) \sim \mathcal{GP}(0,\Sigma)$. The IDE formulation in Eq.~\ref{DiscreteTimeModel} was solved using a numerical method described in Appendix~\ref{Space Discretization}. To generate iEEG/LFP like data we used the output function 
\begin{equation}
	y_{k}\left(r_n\right) = \int_{\Omega}{m\left(|r_n-r'|\right)v_{k}\left(r\right)dr'}
\end{equation}
where $r_n$ defines the location of the sensors in the field, $k$ is the time of sampling instances corresponding to an acquisition system and $m(|r-r'|)$ is the output kernel that governs the geometry of pick-up range of sensors where
\begin{equation}
	m\left(|r-r'|\right) = \exp{\left(-\frac{\left(r_x - r_{x_n}\right)^2}{\left(2\sigma_m\right)^2}-\frac{\left(r_y - r_{y_n}\right)^2}{\left(2\sigma_m\right)^2}\right)}.
\end{equation}

\section{Discretized Model for Simulation}
We begin the derivation of the state-space model with the continuous time, continuous space IDE neural field model. 
Spatio-temporal interactions can be described by the continuous space continuous time stochastic differential equation
\begin{equation}\label{differential eq}	
dv\left( r,t \right) =  \left( -\alpha v\left( r,t \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t \right)} \right)dr'} \right) dt  + d e\left( r,t \right).
\end{equation}
To simplify the notation, but without loss of generality, \ref{FinalForm1} can be rewritten as
\begin{equation}\label{differential simplified}
	dv\left( r,t \right) = q\left(v\left(r,t\right) \right) dt  + d e\left( r,t \right)
\end{equation}
where $q\left(v\left(r,t\right)\right)$ is the term insides the brackets in Eq.~\ref{differential eq}. A one-step Euler method was used to approximate the nonlinear stochastic IDE to a discrete time form. 
\begin{equation}\label{euler}
	v\left(r,t\right) =  v(r,t_0) + \int_{t_0}^{t} q\left(v(r,\tau) \right) d\tau  + \int_{t_0}^{t} d e\left(r,\tau\right)
\end{equation}
Using an Euler approximation,~\ref{euler} can be approximated at each sampling time by
\begin{equation}\label{euler approx 1}
	v\left(r,t_{k+1} \right) =  v(r,t_k) + q\left(v(r,t_k)\right) (t_{k+1} - t_k) + (e(r,t_{k+1}) - e(r,t_k))
\end{equation}
which can be rewritten as
\begin{equation}\label{euler approx 2}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta q\left(v(r,t_k)\right) + e'(r,t_k)
\end{equation}
where $e'(r,t_k)$ is a zero mean Gaussian random variable with covariance $\delta\Sigma_e$. (Note that, because of the way the noise is defined in \ref{differential eq} it is not $\delta^2$ but $\delta$). Expanding $q\left(v(r,t_k)\right)$ we have
\begin{equation}\label{discrete time model}
	v\left(r,t_k + \delta \right) =  v(r,t_k) + \delta  \left( -\alpha v\left( r,t_k \right) + \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} \right) + e'(r,t_k)
\end{equation}
which can be simplified to
\begin{equation}\label{discrete time model2}
	v\left(r,t_k + \delta \right) =  (1 - \delta \alpha) v(r,t_k) + \delta \int_\Omega  {w\left( r,r' \right)f\left( {v\left( r',t_k \right)} \right)dr'} + e'(r,t_k)
\end{equation}
Now we define the spatial aspect of the model on a regular square $i,j$ grid of neural masses, where the spatial step size $\Delta \mathbf{r}_i = \Delta \mathbf{r}_j = \Delta $ giving
\begin{equation}\label{discrete space}
	v\left(r_{ij},t_k + \delta \right) =  \lambda v(r_{ij},t_k) + \delta \Delta^2 \sum_{i=-I}^{i=I}{\sum_{j=-J}^{j=J}{\tilde{w}\left( r_{ij},r_{ij}' \right)f\left( v\left( r_{ij}',t_k \right) \right)} }+ e'(r_{i,j},t_k).
\end{equation}
where $\tilde{w}(r_{ij},r_{ij}')$ discretized connectivity kernel and $\lambda = 1 - \delta \alpha$. In Eq.~\ref{discrete space} we have used periodic boundary conditions to alleviate problems associated with edge effects. The connectivity kernel is in the form of a modified Mexican hat, and is described by the sum of three weighted Gaussian functions as
\begin{equation}
	\tilde{w}(r_{ij},r_{ij}') = \mathbf{\theta}^T e^{\frac{\left(\left(r_i'-r_i\right)\Delta\right)^2}{(2\mathbf{\sigma})^2} + \frac{\left(\left(r_j'-r_j\right)\Delta\right)^2}{(2\mathbf{\sigma})^2}}
\end{equation}
where
\begin{equation}
 \mathbf{\theta}=\begin{bmatrix} \mathbf{\theta}_{1}\\\mathbf{\theta}_{2}\\\mathbf \theta_{3}\end{bmatrix},\quad \mathbf{\sigma}=\begin{bmatrix}\mathbf{\sigma}_1\\\mathbf{\sigma}_2\\ \mathbf{\sigma}_3\end{bmatrix}.
\end{equation}
The first Gaussian function (parametrized by $\mathbf{\theta}_1, \mathbf{\sigma}_1$) describes local excitation, the second describes lateral inhibition and the third describes medium range excitation.

To generate data will approximate either local field potential recordings or intracranial EEG we have used the observation equation
\begin{equation}\label{ObservationEquation}
    y(r_{i_n j_n},t_l) = \Delta^2\sum_{i=1}^{i=I}\sum_{j=1}^{j=J}  m( r_{i_n j_n},r_{ij}' ) v( r_{ij}',t_l )  + \varepsilon ( r_{i_n j_n},t_l ),
\end{equation}
where $\varepsilon( n )\sim\mathcal{N}(0,\Sigma)$, $n$ indexes the iEEG electrodes, and $m(r_n,r)$ is the observation kernel relating the neural field to measurements accounting for electrodes characteristics is described by
\begin{equation}
	m( r_{i_n j_n},r_{ij}' ) = e^{\frac{\left(\left(r_i'-r_{i_n}\right)\Delta\right)^2}{\left(2\sigma_4\right)^2} + \frac{\left(\left(r_j'-r_{j_n}\right)\Delta\right)^2}{\left(2\sigma_4\right)^2}}.
\end{equation}
In Eq.~\ref{ObservationEquation} we have indexed time by $t_l$ corresponding to the sampling rate of a data acquisition system, where the model is simulated on a much fine time-scale of $t_k$. The sensor positions correspond to positions $i_n,j_n$ and are space on a square grid spaced xx units apart, yy units from the edge of the field.

@@

\begin{tabular}{ccc}
\hline
Parameter & Value & Reference\\
\hline
$\alpha$ & $100$ s$^{-1}$ &\\
\hline
$\nu_0$ & $1$ s$^{-1}$ &\\
\hline
$v_0$ & $6$ mV &\\
\hline
$\beta$ & $0.56$ spikes/mV$^{-1}$ & \\
\hline
$\delta$ & $1\times10^{-4}$ s &\\
\hline
$\Delta$ & $1$ &\\
\hline
$\theta_1$ &  &\\
\hline
$\theta_2$ &  &\\
\hline
$\theta_3$ &  &\\
\hline
$\sigma_1$ &  &\\
\hline
$\sigma_2$ &  &\\
\hline
$\sigma_3$ &  &\\
\hline
$\sigma_4$ &  &\\
\hline
\end{tabular}

@@

In order to implement standard estimation techniques we propose a decomposition of the field using a set of Gaussian basis functions. Decomposition allows a continuous field to be represented by a finite dimensional state vector. This allows for the application of standard nonlinear, non-Gaussian state estimation methods such as the unscented Kalman filter. The abbreviate the notation we use the subscript $t$ to index the current time step and $t+1$ for next time step. The field decomposition is described by
\begin{equation}\label{DefFieldDecomp}
v_t\left( \mathbf{r} \right) \approx \phi ^T\left( \mathbf{r} \right) x_t.
\end{equation}
where $\phi(\mathbf{r})$ is a vector of Gaussian basis functions that are scaled by the state vector, $x_t$. The connectivity kernel can all be decomposed into a set of basis functions
\begin{equation}\label{DefKernelDecomp}
\tilde{w}( r_{ij}, r_{ij}' ) = \theta^T\psi\left(r_{ij},r_{ij}'\right)
\end{equation}
Will will assume we know the parametric form of the basis functions, where the parameter $\theta$ is unknown. Each connectivity basis function can be considered a layer in the WC model, representing short range excitation, surround inhibition and mid-range excitation. Making substitutions of~\ref{DefFieldDecomp} and~\ref{DefKernelDecomp} into~\ref{discrete space} we get

@@

\begin{equation}\label{reduced discrete model}
	\phi ^T\left( r_{ij} \right)x_{t+1} =  \lambda \phi ^T\left( r_{ij} \right)x_{t+1} + \delta \Delta^2 \theta^T \sum_{i=-I}^{i=I}{\sum_{j=-J}^{j=J}{\psi\left(r_{ij},r_{ij}'\right)f\left( \phi ^T\left(r_{ij} \right)x_{t+1} \right)} }+ e'(r_{i,j},t_k).
\end{equation}

\begin{equation}\label{reduced continuous model}
\phi ^T\left( \mathbf{r} \right)x_{t+1} = T_s\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\mathbf{r'}}} \right)f\left( {{\phi ^T}\left( {{\mathbf{r'}}} \right){x_t}} \right)d{\mathbf{r'}}}  - \lambda\phi^T(\mathbf{r})x_t + T_s{e_t}\left( {\mathbf{r}} \right).
\end{equation}
To isolate the state we cross multiply by $\phi(\mathbf{r})$ and integrate over $\mathbf{r}$
\begin{equation}\label{Poo}
\int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}{x_{t+1}}}  = T_s\int_\Omega  {\phi ( \mathbf{r} )\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}}  - \lambda\int_\Omega \phi(\mathbf{r})\phi^T(\mathbf{r})d\mathbf{r} x_t  + T_s\int_\Omega  {\phi \left( {\mathbf{r}} \right)e_t\left( {\mathbf{r}} \right)d{\mathbf{r}}}.
\end{equation}
Now we define
\begin{equation}\label{DefGamma}
\Gamma  = \int_\Omega  {\phi \left( {\bf{r}} \right){\phi ^T}\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
Substituting equation~\ref{DefGamma} into \ref{Poo} and rearranging gives
\begin{equation}\label{SS Model}
{x_{t+1}} = T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)\theta^T\int_\Omega  {\psi \left( {{\bf{r}} , {\bf{r'}}} \right)f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)d{\bf{r'}}} d{\bf{r}}} - \lambda x_t + T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)e_t\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
We can rearrange to get
\begin{equation}\label{Homogeneous SS Model}
	x_{t+1} = T_s\Gamma ^{ - 1}\int_\Omega  {\phi \left( {\bf{r}} \right)\int_\Omega  {f\left( {{\phi ^T}\left( {{\bf{r'}}} \right){x_t}} \right)\psi^T \left( {{\bf{r}} - {\bf{r'}}} \right)d{\bf{r'}}} d{\bf{r}}}\theta - \lambda x_t + T_s{\Gamma ^{ - 1}}\int_\Omega  {\phi \left( {\bf{r}} \right)e_t\left( {\bf{r}} \right)d{\bf{r}}}.
\end{equation}
The disturbance term can be rewritten as 
\begin{equation}
	\Gamma ^{ - 1}\int_\Omega  {\phi \left( \bf{r} \right)e_t\left( \bf{r} \right)d\bf{r}} = \int_\Omega  {\phi \left( \bf{r} \right)\tilde{e}_t\left( \bf{r} \right)d\bf{r}},
\end{equation}
where $\tilde{e}_t(r)~N(0,R'R)$ and $R'$ is the Cholesky decomposition of $\Gamma^{-1}$. This provides the final state-space model
\begin{equation}\label{AbbrevSSModel}
x_{t + 1} = T_sq(x_t)\theta -\lambda x_t + T_s\int_\Omega  {\phi ( \mathbf{r} )\tilde{e}_t( \mathbf{r} )d\mathbf{r}}
\end{equation}
with observation equation
\begin{equation}\label{ObservationEquation}
    y_t( k ) = \int_\Omega  m( k,\mathbf{r} )\phi^T( \mathbf{r} ) d\mathbf{r}x_t  + \varepsilon ( k ),
\end{equation}

% \subsection{Discrete Space For Simulation}
% 
% In this section we describe a discrete time, discrete space model that we are using for the simulations. Now we define the spatial aspect of the model on a regular square $i,j$ grid of neural masses, where the spatial step size $\Delta \mathbf{r}_i = \Delta \mathbf{r}_j = \Delta \mathbf{r}$ giving
% \begin{equation}\label{DiscreteSpaceModelij}
% 	x_{t+1} = T_s\Gamma ^{ - 1}\sum_i \sum_j  {\phi \left( \mathbf{r} \right)\sum_i\sum_j  {f\left( \phi^T \left( \mathbf{r'} \right) x_t \right)\psi^T \left( \mathbf{r} - \bf{r'} \right)\Delta\mathbf{r'}^2} \Delta\mathbf{r}^2} \theta - \lambda x_t + T_s\sum_i\sum_j  {\phi \left( {\bf{r}} \right)\tilde{e}_t\left( \bf{r} \right)\Delta\mathbf{r}^2}.
% \end{equation}
\section{Estimation of the Nonlinear Homogeneous IDE Neural Field Model}
